---
title: "Creating Landscapes in gen3sis"
date: "`r format(Sys.time(),  '%d.%m.%Y')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating Landscapes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

To create your own landscape, you can start from anywere. For this exercise we will recreate the simple island landscape that we used in the introduction.

```{r setup}
library(gen3sis)
```

## Create your rasters

```{r eval=FALSE}
library(raster)
library(landscapetools)
library(NLMR)

setwd(getwd())

##Experiment 1: Island dynamics with area and heterogeneity dynamics
#Spatial dimensions:
Ncol<-31
Nrow<-31
#Temporal extent:
Ntimesteps<-140
#Creating an array:
My_Dynamic_Landscape<-array(NA,dim=c(Nrow,Ncol,Ntimesteps))
My_Dynamic_Landscape_dataframe<-matrix(NA,nrow=Nrow*Ncol, ncol=length(c(Ncol,Nrow))+Ntimesteps)
My_Dynamic_Landscape2<-array(NA,dim=c(Nrow,Ncol,Ntimesteps))
My_Dynamic_Landscape_dataframe2<-matrix(NA,nrow=Nrow*Ncol, ncol=length(c(Ncol,Nrow))+Ntimesteps)
stringtimestepsnames<-vector(mode="character",length=Ntimesteps)
#My_Dynamic_Landscape[,1:2,] <- 20   #Error in file_path_sans_ext(basename(config_file)) : could not find function "file_path_sans_ext"
#My_Dynamic_Landscape2[,1:2,] <- 500

#Looping over the time steps and creating a landmass dynamics, 
#while saving temperature values on it:
for (timestep in 1:Ntimesteps){
  counting<-1
  stringtimestepsnames[timestep]<-paste("X",timestep,"", sep="")
  for (y in 1:Nrow){
    for (x in 1:Ncol){
      if((timestep<=10)||(timestep>120 && timestep<=140)){
        if((y>=15 && y<=16)&&(x>=15 && x<=16)) {
          My_Dynamic_Landscape[x,y,timestep] <- rnorm(1,20,0.5) 
          My_Dynamic_Landscape2[x,y,timestep] <- rnorm(1,500,50) 
        }
      }
      if((timestep>10 && timestep<=20)||(timestep>100 && timestep<=120)){
        if((y>=15 && y<=17)&&(x>=15 && x<=17)) My_Dynamic_Landscape[x,y,timestep] <- rnorm(1,20,0.5) 
        if(y==16 && x==16) My_Dynamic_Landscape[x,y,timestep] <- rnorm(1,19,0.5) 
        
        if((y==15)&&(x>=15 && x<=17)) My_Dynamic_Landscape2[x,y,timestep] <- rnorm(1,500,50) 
        if((y==16)&&(x>=15 && x<=17)) My_Dynamic_Landscape2[x,y,timestep] <- rnorm(1,600,50) 
        if((y==17)&&(x>=15 && x<=17)) My_Dynamic_Landscape2[x,y,timestep] <- rnorm(1,700,50)   
      }
      if((timestep>20 && timestep<=30)||(timestep>80 && timestep<=100)){
        if((y>=14 && y<=18)&&(x>=14 && x<=18)) My_Dynamic_Landscape[x,y,timestep] <- rnorm(1,20,0.5) 
        if((y>=15 && y<=17)&&(x>=15 && x<=17)) My_Dynamic_Landscape[x,y,timestep] <- rnorm(1,19,0.5) 
        if(y==16 && x==16) My_Dynamic_Landscape[x,y,timestep] <- rnorm(1,18,0.5) 
        
        if((y==14)&&(x>=14 && x<=18)) My_Dynamic_Landscape2[x,y,timestep] <- rnorm(1,500,50) 
        if((y==15)&&(x>=14 && x<=18)) My_Dynamic_Landscape2[x,y,timestep] <- rnorm(1,600,50) 
        if((y==16)&&(x>=14 && x<=18)) My_Dynamic_Landscape2[x,y,timestep] <- rnorm(1,700,50) 
        if((y==17)&&(x>=14 && x<=18)) My_Dynamic_Landscape2[x,y,timestep] <- rnorm(1,800,50) 
        if((y==18)&&(x>=14 && x<=18)) My_Dynamic_Landscape2[x,y,timestep] <- rnorm(1,900,50)         
      }
      if((timestep>30 && timestep<=40)||(timestep>60 && timestep<=80)){
        if((y>=13 && y<=19)&&(x>=13 && x<=19)) My_Dynamic_Landscape[x,y,timestep] <- rnorm(1,20,0.5) 
        if((y>=14 && y<=18)&&(x>=14 && x<=18)) My_Dynamic_Landscape[x,y,timestep] <- rnorm(1,19,0.5) 
        if((y>=15 && y<=17)&&(x>=15 && x<=17)) My_Dynamic_Landscape[x,y,timestep] <- rnorm(1,18,0.5) 
        if(y==16 && x==16) My_Dynamic_Landscape[x,y,timestep] <- rnorm(1,17,0.5) 
        
        if((y==13)&&(x>=13 && x<=19)) My_Dynamic_Landscape2[x,y,timestep] <- rnorm(1,500,50) 
        if((y==14)&&(x>=13 && x<=19)) My_Dynamic_Landscape2[x,y,timestep] <- rnorm(1,600,50) 
        if((y==15)&&(x>=13 && x<=19)) My_Dynamic_Landscape2[x,y,timestep] <- rnorm(1,700,50) 
        if((y==16)&&(x>=13 && x<=19)) My_Dynamic_Landscape2[x,y,timestep] <- rnorm(1,800,50) 
        if((y==17)&&(x>=13 && x<=19)) My_Dynamic_Landscape2[x,y,timestep] <- rnorm(1,900,50) 
        if((y==18)&&(x>=13 && x<=19)) My_Dynamic_Landscape2[x,y,timestep] <- rnorm(1,1000,50)  
        if((y==19)&&(x>=13 && x<=19)) My_Dynamic_Landscape2[x,y,timestep] <- rnorm(1,1100,50)           
      }
      if(timestep>40 && timestep<=60){
        if((y>=12 && y<=20)&&(x>=12 && x<=20)) My_Dynamic_Landscape[x,y,timestep] <- rnorm(1,20,0.5) 
        if((y>=13 && y<=19)&&(x>=13 && x<=19)) My_Dynamic_Landscape[x,y,timestep] <- rnorm(1,19,0.5) 
        if((y>=14 && y<=18)&&(x>=14 && x<=18)) My_Dynamic_Landscape[x,y,timestep] <- rnorm(1,18,0.5) 
        if((y>=15 && y<=17)&&(x>=15 && x<=17)) My_Dynamic_Landscape[x,y,timestep] <- rnorm(1,17,0.5) 
        if(y==16 && x==16) My_Dynamic_Landscape[x,y,timestep] <- rnorm(1,16,0.5) 
        
        if((y==12)&&(x>=12 && x<=20)) My_Dynamic_Landscape2[x,y,timestep] <- rnorm(1,500,50) 
        if((y==13)&&(x>=12 && x<=20)) My_Dynamic_Landscape2[x,y,timestep] <- rnorm(1,600,50) 
        if((y==14)&&(x>=12 && x<=20)) My_Dynamic_Landscape2[x,y,timestep] <- rnorm(1,700,50) 
        if((y==15)&&(x>=12 && x<=20)) My_Dynamic_Landscape2[x,y,timestep] <- rnorm(1,800,50) 
        if((y==16)&&(x>=12 && x<=20)) My_Dynamic_Landscape2[x,y,timestep] <- rnorm(1,900,50) 
        if((y==17)&&(x>=12 && x<=20)) My_Dynamic_Landscape2[x,y,timestep] <- rnorm(1,1000,50)  
        if((y==18)&&(x>=12 && x<=20)) My_Dynamic_Landscape2[x,y,timestep] <- rnorm(1,1100,50)  
        if((y==19)&&(x>=12 && x<=20)) My_Dynamic_Landscape2[x,y,timestep] <- rnorm(1,1200,50)  
        if((y==20)&&(x>=12 && x<=20)) My_Dynamic_Landscape2[x,y,timestep] <- rnorm(1,1300,50)  
      }

      if(timestep==1){
        My_Dynamic_Landscape_dataframe[counting,1]<-x
        My_Dynamic_Landscape_dataframe[counting,2]<-y
        My_Dynamic_Landscape_dataframe2[counting,1]<-x
        My_Dynamic_Landscape_dataframe2[counting,2]<-y
      }
      My_Dynamic_Landscape_dataframe[counting,2+timestep]<-My_Dynamic_Landscape[x,y,timestep]
      My_Dynamic_Landscape_dataframe2[counting,2+timestep]<-My_Dynamic_Landscape2[x,y,timestep]
      counting<-counting+1
    }
  }
  My_Dynamic_Landscape_dataframe_suitable<-My_Dynamic_Landscape_dataframe[which(My_Dynamic_Landscape_dataframe[,2+timestep]>0),]
  if(is.null(dim(My_Dynamic_Landscape_dataframe_suitable))) dist_matrx<-as.matrix(dist(My_Dynamic_Landscape_dataframe_suitable[1:2], method = "euclidean"))
  else dist_matrx<-as.matrix(dist(My_Dynamic_Landscape_dataframe_suitable[,1:2], method = "euclidean"))
  ##saving the dist_matrix into .rds format to be used as input
  saveRDS(dist_matrx, paste(getwd(),"/input/my_experiment/distance_matrices/geo_dist_m_ti_t_",(Ntimesteps - timestep),".rds", sep=""))
  
  #Sys.sleep(0.4)
  #image(My_Dynamic_Landscape[,,timestep])
}

which(!is.na(My_Dynamic_Landscape_dataframe[,timestep]))

My_Dynamic_Landscape_dataframe<-as.data.frame(My_Dynamic_Landscape_dataframe)
colnames(My_Dynamic_Landscape_dataframe)<- c("x","y",stringtimestepsnames)
My_Dynamic_Landscape_dataframe2<-as.data.frame(My_Dynamic_Landscape_dataframe2)
colnames(My_Dynamic_Landscape_dataframe2)<- c("x","y",stringtimestepsnames)
is.data.frame(My_Dynamic_Landscape_dataframe)
is.list(My_Dynamic_Landscape_dataframe)
all_geo_hab <- list("temp" = My_Dynamic_Landscape_dataframe,"prec"=My_Dynamic_Landscape_dataframe2) #This is important, otherwise the objects in the list won't have a name
##saving the list of data frames into .rds format to be used as input
saveRDS(all_geo_hab, paste(getwd(),"/input/my_experiment/all_geo_hab.rds", sep=""))



##Experiment 2: Island dynamics with only area dynamics
#Spatial dimensions:
Ncol<-31
Nrow<-31
#Temporal extent:
Ntimesteps<-140
#Creating an array:
My_Dynamic_Landscape<-array(NA,dim=c(Nrow,Ncol,Ntimesteps))
My_Dynamic_Landscape_dataframe<-matrix(NA,nrow=Nrow*Ncol, ncol=length(c(Ncol,Nrow))+Ntimesteps)
My_Dynamic_Landscape2<-array(NA,dim=c(Nrow,Ncol,Ntimesteps))
My_Dynamic_Landscape_dataframe2<-matrix(NA,nrow=Nrow*Ncol, ncol=length(c(Ncol,Nrow))+Ntimesteps)
stringtimestepsnames<-vector(mode="character",length=Ntimesteps)
#My_Dynamic_Landscape[,1:2,] <- 20   #Error in file_path_sans_ext(basename(config_file)) : could not find function "file_path_sans_ext"
#My_Dynamic_Landscape2[,1:2,] <- 500

#Looping over the time steps and creating a landmass dynamics, 
#while saving temperature values on it:
for (timestep in 1:Ntimesteps){
  counting<-1
  stringtimestepsnames[timestep]<-paste("X",timestep,"", sep="")
  for (y in 1:Nrow){
    for (x in 1:Ncol){
      if((timestep<=10)||(timestep>120 && timestep<=140)){
        if((y>=15 && y<=16)&&(x>=15 && x<=16)) {
          My_Dynamic_Landscape[x,y,timestep] <- rnorm(1,20,0.5) 
          My_Dynamic_Landscape2[x,y,timestep] <- rnorm(1,500,50) 
        }
      }
      if((timestep>10 && timestep<=20)||(timestep>100 && timestep<=120)){
        if((y>=15 && y<=17)&&(x>=15 && x<=17)) {
          My_Dynamic_Landscape[x,y,timestep] <- rnorm(1,20,0.5) 
          My_Dynamic_Landscape2[x,y,timestep] <- rnorm(1,500,50)
        }
      }
      if((timestep>20 && timestep<=30)||(timestep>80 && timestep<=100)){
        if((y>=14 && y<=18)&&(x>=14 && x<=18)) {
          My_Dynamic_Landscape[x,y,timestep] <- rnorm(1,20,0.5) 
          My_Dynamic_Landscape2[x,y,timestep] <- rnorm(1,500,50) 
        }
      }
      if((timestep>30 && timestep<=40)||(timestep>60 && timestep<=80)){
        if((y>=13 && y<=19)&&(x>=13 && x<=19)) {
          My_Dynamic_Landscape[x,y,timestep] <- rnorm(1,20,0.5)
          My_Dynamic_Landscape2[x,y,timestep] <- rnorm(1,500,50)
        }
      }
      if(timestep>40 && timestep<=60){
        if((y>=12 && y<=20)&&(x>=12 && x<=20)) {
          My_Dynamic_Landscape[x,y,timestep] <- rnorm(1,20,0.5)
          My_Dynamic_Landscape2[x,y,timestep] <- rnorm(1,500,50)  
        }
      }
      if(timestep==1){
        My_Dynamic_Landscape_dataframe[counting,1]<-x
        My_Dynamic_Landscape_dataframe[counting,2]<-y
        My_Dynamic_Landscape_dataframe2[counting,1]<-x
        My_Dynamic_Landscape_dataframe2[counting,2]<-y
      }
      My_Dynamic_Landscape_dataframe[counting,2+timestep]<-My_Dynamic_Landscape[x,y,timestep]
      My_Dynamic_Landscape_dataframe2[counting,2+timestep]<-My_Dynamic_Landscape2[x,y,timestep]
      counting<-counting+1
    }
  }
  My_Dynamic_Landscape_dataframe_suitable<-My_Dynamic_Landscape_dataframe[which(My_Dynamic_Landscape_dataframe[,2+timestep]>0),]
  if(is.null(dim(My_Dynamic_Landscape_dataframe_suitable))) dist_matrx<-as.matrix(dist(My_Dynamic_Landscape_dataframe_suitable[1:2], method = "euclidean"))
  else dist_matrx<-as.matrix(dist(My_Dynamic_Landscape_dataframe_suitable[,1:2], method = "euclidean"))
  ##saving the dist_matrix into .rds format to be used as input
  saveRDS(dist_matrx, paste(getwd(),"/input/my_experiment_2/distance_matrices/geo_dist_m_ti_t_",(Ntimesteps - timestep),".rds", sep=""))
  
  #Sys.sleep(0.4)
  #image(My_Dynamic_Landscape[,,timestep])
}

which(!is.na(My_Dynamic_Landscape_dataframe[,timestep]))

My_Dynamic_Landscape_dataframe<-as.data.frame(My_Dynamic_Landscape_dataframe)
colnames(My_Dynamic_Landscape_dataframe)<- c("x","y",stringtimestepsnames)
My_Dynamic_Landscape_dataframe2<-as.data.frame(My_Dynamic_Landscape_dataframe2)
colnames(My_Dynamic_Landscape_dataframe2)<- c("x","y",stringtimestepsnames)
is.data.frame(My_Dynamic_Landscape_dataframe)
is.list(My_Dynamic_Landscape_dataframe)
all_geo_hab <- list("temp" = My_Dynamic_Landscape_dataframe,"prec"=My_Dynamic_Landscape_dataframe2) #This is important, otherwise the objects in the list won't have a name
##saving the list of data frames into .rds format to be used as input
saveRDS(all_geo_hab, paste(getwd(),"/input/my_experiment_2/all_geo_hab.rds", sep=""))

```
## Add environmental characteristics

## Add a resistance (how connected and reachebel things are)

The distance matrix for each time step between either: (i) full distance, containing all habitable cells in the landscape (faster computing speed, larger storage required), or (ii) local distance, containing a list of distance matrices of each cell with its 16 neighbors (slower computing speed, smaller storage required). Local distances matrix are built at each timestep with equal dimension to the full distances during a simulation run.The real geographic coordinate system can be provided which allows for the global spherical distance correction. For real landscapes, the distance unit is in kilometers (km). In the case of hypothetical landscapes, geographic distance is measured in units of grid cells.The model uses serialized R files (rds) format for all inputs. The functions for processing the input data in the required format are available in the package. The package offers two options to create distance matrices: one can either compute a (i) full distance matrix at each time step, or compute a (ii) local distance matrix of all cells containing the distances between neighbouring cells. The former option will lead to faster simulations, while the latter uses less storage space. Once the data is processed, the simulation can be started with the function ‘run_simulation’.’ 