---
title: "Introducing gen3sis"
date: "`r format(Sys.time(),  '%d.%m.%Y')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introducing gen3sis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
Hello and wellcome! In this vignette you will be introduced to gen3sis, an engine for simulating various spatial eco-evolutionary models. We will create an virtual planet, 'biodiversify' it and examine some brand new virtual species.

Thus we will: 

1. **Start** gen3sis package and check if we are set with all the necessary input data.
2. **Run** one simple example simulation
3. **Visualize** the output
4. **Analize** the output

## 1. Start
First of all, we need to load the gen3sis package 
```{r setup}
library(gen3sis)
```
Perfect! From now on, if you run into problems, use *>??gen3sis* for the full help documentation

Let's check the version of the gen3sis package we are using

```{r, eval=FALSE}
print(paste("GEN3SIS version:", packageVersion("gen3sis")))

```

All gen3sis simulations need a **landscape** and a **configuration** object to refer to. The **landscape** is the environment of the virtual planet (e.g. temperature and precipitation data that changes spatialy and temporaly). The **configuration** are the rules of our virtual planet, here is were we define all the ecological and evolutionary 
Here we will use a published small theoretical planet example[^1]. 

Download the [landscape](https://github.com/project-Gen3sis/Simulations/tree/master/landscape/CaseStudy1) and the [configuration](https://github.com/project-Gen3sis/Simulations/tree/master/config/CaseStudy1) and store these in a new folder wherever you would like to run the experiment.Like this:

|>MY_EXPERIMENT_FOLDER     |
|:-------------------------|
|&nbsp;&nbsp;&nbsp;>landscape|
|&nbsp;&nbsp;&nbsp;>config|

Now that we have the files on the proper folders, we will prepare the landscape.

```{r}
##TODO
# create_input(landscapes = "EXPERIMENT_X/landscape/landscapes.rds", output_directory = )
```

**NOTE** that the landscape contain several files. The .rds file has the landscape changing over time and the distances folder stores some data to inform the landscape is connected.

Done? Great now you have downloaded all the needed files to run a simulation. 


[^1]: Hagen et al. 2020 Methods paper in preparation.


## 2. Run

To launch our simulation use 

```{r, eval=FALSE}
run_simulation(config_file = "./EXPERIMENT_X/config/config", input_directory = "./EXPERIMENT_X/landscape")
```

## 3. Visualize

## 4. Analize



```{r}
 format(Sys.time(),  "%d.%m.%Y")
```

```{r eval=FALSE}
run_simulation(X,y,z)
```